Flexible Probabilistic Forecasting with Distributional Trees and Forests

### 2018-06-13 pkg/disttree/inst/eeecon_workshop/slides.Rnw
presentation at the eeecon workshop (2018-06-27)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2190 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: dff0d574fef171aacc7d4d1fffb6f1d98f87e649

\title{Flexible Probabilistic Forecasting with Distributional Trees and Forests}
\author{Lisa Schlosser}
%\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
%\URL{https://R-Forge.R-project.org/projects/partykit/}
\headerimage{3}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\graphicspath{{plots/}}

@@

library("ggplot2")
theme_set(theme_bw(base_size = 18))
library("disttree")
#library("partykitR1")
library("gamlss")
library("lattice")
library("crch")
library("latex2exp")
library("parallel")
library("RainTyrol")
library("gamlss.cens")
gen.cens(NO, type = "left")

@@

\vspace{-3.5cm}
\visible<6->{
\textbf{Application/Motivation:}
\begin{itemize}
\item Prediction.
\item Caption of dependencies and interactions.
\end{itemize}
}
\vspace{0.5cm}
\visible<7->{
\textbf{Two main approaches / ``cultures'':}
\begin{itemize}
\item Data modeling.
\item Algorithmic modeling.
\end{itemize}
}
\end{frame}

@@

\begin{frame}
\frametitle{Goals}
\vspace{-0.4cm}
\begin{minipage}{0.7\textwidth}
\textbf{Available:} \\
\vspace{-0.5 cm}
\begin{itemize}
  \item Distributional regression:\\
  Probabilistic forecast.
  \item Regression Trees:\\
  Automatic detection of abrupt changes,
  and selection of variables/interactions.
  \item Random Forests:\\
  Smoothing and regularization.
\end{itemize}

@@

$\Rightarrow$ PhD-Thesis

@@

%{\small 
(what I've been working on the last 2 years)
}
\end{minipage}
<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
nobs <- 200
x_g <- c(1:nobs)/nobs
x_g <- 2*(x_g-0.5)
ytrue_g <- x_g^3
var_g <- exp(-(2*x_g)^2)/2
y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
@
<<motivation_regtree_syn, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(74)
nobs <- 170
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- var <- var_up <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<0.2) 4 else 4+5*plogis(kappa*(3*(x[i]-0.33)-1))
ytrue_up <- rep.int(15, length(x))
for(i in 1:nobs) var_up[i] <- if(x[i]<0.15) 3 else (if(x[i]<0.35) 2 else (if(x[i]<0.5) 2 else (if(x[i]<0.7) 1 else (if(x[i]<0.8) 1.5 else (if(x[i]<0.9) 2 else (2.5))))))
for(i in 1:nobs) var[i] <- if(x[i]<0.15) 4 else (if(x[i]<0.2) 3 else (if(x[i]<0.25) 2 else (if(x[i]<0.4) 1 else (if(x[i]<0.8) 2 else (if(x[i]<0.9) 2.5 else (3))))))
y <- ytrue + rnorm(nobs,0, sqrt(var))
y_up <- ytrue_up + rnorm(nobs,0, sqrt(var_up))
for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
ytree_up <- rep.int(15, length(x))
for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
@
\begin{minipage}{0.25\textwidth}
\vspace{0.4cm}
<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
  col = pallight["forest"], border = "transparent")
lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
points(x_g, y_g, col = "slategray", pch = 19)
box(lwd = 5)
@

@@

plot(tree, tnex = 1.7, drop = TRUE,
     terminal_panel = node_density,
     tp_args = list(fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\visible<1->{
R-package \code{disttree}
}
\end{center}
\end{minipage}
\begin{minipage}{0.21\textwidth}

@@

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
Estimate model parameters $\hat{\theta}$.
\item<6-> Evaluate goodness of fit \\
(for each parameter and each observation).
\item<8-> Choose covariate $X$ with strongest influence on goodness of fit of $\mathcal{D}(Y; \hat{\theta})$ as split variable.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.25\textwidth}
\vspace{-1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density1) at (0.8,2)
    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
}
\visible<7-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1);
\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2);
}
\visible<9-10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
}
\visible<11->{
\draw[-, line width=0.8pt] (0.7, 1.4) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (0.9, 1.4) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12->{
\node[inner sep=0pt] (density2) at (-0.1,-0.25)
    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
}
\visible<12->{
\node[inner sep=0pt] (density3) at (1.7,-0.25)
    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
\begin{adjustwidth}{-0.0em}{-3em}
\begin{enumerate}
\setcounter{enumi}{3}
\item<8-> Find the split point $p$ which leads to the highest improvement.
\item<10-> Repeat steps 1--4 recursively in the subgroups
    until some stopping criterion is met.
\end{enumerate}
\end{adjustwidth}
\vspace{0.4cm}
\visible<13->{
{\bf Forest:} Ensemble of trees.
\begin{itemize}
\item Varying subsamples.
\item Combine / average over trees.
\end{itemize}
}
\end{frame}

@@

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
  \url{http://arxiv.org/abs/1804.02921}

>>>>

\title{Flexible Probabilistic Forecasting with Distributional Trees and Forests}
\author{Lisa Schlosser, Torsten Hothorn, Reto Stauffer, Achim Zeileis}
%\author{Lisa Schlosser, Torsten Hothorn,\\ Heidi Seibold, Reto Stauffer, Achim Zeileis}
\setbeamerfont{url}{size*={11.5pt}{13pt},series=\mdseries}
\URL{https://R-Forge.R-project.org/projects/partykit/}
\headerimage{3}

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
\usepackage{changepage}
\usepackage{amsmath,tikz}
\usepackage{calc}
\usepackage{graphicx}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap,trees,tikzmark,decorations.pathreplacing}
\graphicspath{{plots/}}
\newcommand{\argmax}{\operatorname{argmax}\displaylimits}

@@

# setwd("~/svn/partykit/pkg/disttree/inst/eeecon_workshop")
library("ggplot2")
theme_set(theme_bw(base_size = 18))
library("disttree")
#library("partykitR1")
library("gamlss")
library("lattice")
library("crch")
library("latex2exp")
library("parallel")
library("RainTyrol")
library("gamlss.cens")
gen.cens(NO, type = "left")

@@

\vspace{-3.5cm}
\visible<6->{
\textbf{Application/Motivation:}
\begin{itemize}
\item Prediction.
\item Capturing dependencies and interactions.
\end{itemize}
}
\vspace{0.5cm}
\visible<7->{
\textbf{Two main approaches / ``cultures'':}
\begin{itemize}
\item Data modeling.
\item Algorithmic modeling.
\end{itemize}
}
\end{frame}

@@

\begin{frame}
\frametitle{Goals}
\vspace{-0.4cm}
\begin{minipage}{0.7\textwidth}
\textbf{Available:} \\
\vspace{-0.5 cm}
\begin{itemize}
  \item Distributional regression:\\
  Probabilistic forecast.
  \item Regression trees:\\
  Automatic detection of abrupt changes,
  and selection of variables/interactions.
  \item Random forests:\\
  Smoothing and regularization.
\end{itemize}

@@

$\Rightarrow$ Distributional trees and forests

@@

}
\end{minipage}
<<motivation_GAMLSS_syn, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
nobs <- 200
x_g <- c(1:nobs)/nobs
x_g <- 2*(x_g-0.5)
ytrue_g <- x_g^3
var_g <- exp(-(2*x_g)^2)/2
y_g <- ytrue_g + rnorm(nobs, 0, 0.1 + var_g)
@
<<motivation_regtree_syn, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(74)
nobs <- 170
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- ytree2 <- ytree3 <- ytree4 <- ytree_up <- yforest <- var <- var_up <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<0.2) 4 else 4+5*plogis(kappa*(3*(x[i]-0.33)-1))
ytrue_up <- rep.int(15, length(x))
for(i in 1:nobs) var_up[i] <- if(x[i]<0.15) 3 else (if(x[i]<0.35) 2 else (if(x[i]<0.5) 2 else (if(x[i]<0.7) 1 else (if(x[i]<0.8) 1.5 else (if(x[i]<0.9) 2 else (2.5))))))
for(i in 1:nobs) var[i] <- if(x[i]<0.15) 4 else (if(x[i]<0.2) 3 else (if(x[i]<0.25) 2 else (if(x[i]<0.4) 1 else (if(x[i]<0.8) 2 else (if(x[i]<0.9) 2.5 else (3))))))
y <- ytrue + rnorm(nobs,0, sqrt(var))
y_up <- ytrue_up + rnorm(nobs,0, sqrt(var_up))
for(i in 1:nobs) ytree[i] <- if(x[i]<0.7) 4 else 9
ytree_up <- rep.int(15, length(x))
for(i in 1:nobs) yforest[i] <- if(x[i]<0.55) 4 else {if(x[i]<0.6) 5 else {if(x[i]<0.7) 6.5 else {if(x[i]<0.75) 8 else 9}}}
@
\begin{minipage}{0.25\textwidth}
\vspace{0.4cm}
<<plot_motivation_GAMLSS_syn, fig=TRUE, echo=FALSE, width=8.5>>=
par(mar=c(1,0,1,0))
plot(x_g, y_g, xaxt = "n", yaxt = "n", ann = FALSE, type = "n")
polygon(c(x_g, rev(x_g)), c(ytrue_g + 0.1 + var_g, rev(ytrue_g - 0.1 - var_g)),
  col = pallight["forest"], border = "transparent")
lines(x_g, ytrue_g, col = pal["forest"], lwd=10)
points(x_g, y_g, col = "slategray", pch = 19)
box(lwd = 5)
@

@@

plot(tree, tnex = 1.7, drop = TRUE,
     terminal_panel = node_density,
     tp_args = list(fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]))
@
}
\visible<1->{
R~package~\code{disttree}
}
\end{center}
\end{minipage}
\begin{minipage}{0.21\textwidth}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{frame}
%\frametitle{Global Likelihood Estimation}
%\begin{itemize}

%\item Specify a parametric distribution family $F(\cdot;\theta)$ with parameter vector
%  $\theta \in \Theta$ capturing location, scale, shape.

%\item Log-likelihood function:
%  \begin{eqnarray*}
%  \ell(\theta; y) & = & \log(f(y; \theta))
%  \end{eqnarray*}

%\item Estimate $\hat{\bold{\theta}}$ via maximum likelihood
%  based on a learning sample $y_1,\dots, y_n$:
%\[
%\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)
%\]

%% equivalent to solving 
%% $$
%% \sum_{i=1}^n \frac{\partial \ell}{\partial \theta}(\hat{\theta}; y_i) = \sum_{i=1}^n s(y_i, \hat{\theta}) = 0.
%% $$
%\end{itemize}

%\begin{frame}
%\frametitle{Adaptive Local Likelihood Estimation}

%\textbf{Idea:} Capture covariates through adaptive weights.
%\[
%\hat{\theta}(\bold{x}) = \argmax_{\theta \in \Theta} \sum_{i=1}^n w_i(\bold{x}) \cdot \ell%(\theta; y_i).
%\]

%\bigskip

%\textbf{Question:} How to choose weighting function $w_i(\bold{x})$?

%\bigskip

%\textbf{Possible answers:} Based on learning sample $\{(y_i, \bold{x}_i)\}_{i=1, \dots, n}$ and (possibly new) observation $\bold{x}$.
%\begin{itemize}
%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $\bold{x}_i$ are classified into the same subgroup.
%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $\bold{x}_i$ across trees.
%\end{itemize}

%\textbf{Possible answers:} Based on learning sample $y_1, \dots, y_n$ and (possibly new) observation $\bold{x}$.
%\begin{itemize}
%\item \emph{Tree:} $w_i(\bold{x}) \in \{0,1\}$ indicates whether $\bold{x}$ and $y_i$ are classified into the same subgroup.
%\item \emph{Forest:} $w_i(\bold{x}) \in  [0, 1]$ averages the weights for $\bold{x}$ and $y_i$ across trees.
%\end{itemize}

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood\\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{D}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density1) at (0.8, 1.7)
    {\includegraphics[width=0.6\textwidth]{density1.jpeg}};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1);
\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2);
}
\visible<9-10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
}
\visible<11->{
\draw[-, gray, line width=0.5pt] (0.7, 1.18) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (0.9, 1.18) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12->{
\node[inner sep=0pt] (density2) at (-0.1,-0.1)
    {\includegraphics[width=0.5\textwidth]{density2.jpeg}};
}
\visible<12->{
\node[inner sep=0pt] (density3) at (1.7,-0.1)
    {\includegraphics[width=0.5\textwidth]{density3.jpeg}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<8-> Split along the covariate $X$ with strongest association or instability and at breakpoint $p$ with highest improvement \\
in log-likelihood.
\item<10-> Repeat steps 1--3 recursively until some stopping criterion \\
is met, yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.4cm}
\visible<13->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

\begin{frame}
\frametitle{Adaptive Local Likelihood Estimation}

Parameter estimator for a \only<1-3>{global}\only<4->{local} model with learning data \only<1>{$\{y_i\}_{i=1,\ldots,n}$}\only<2->{$\{(y_i,\bold{x}_i)\}_{i=1,\ldots,n}$}:

\[
\hat{\theta}\visible<2->{(\bold{x})} =  \argmax_{\theta \in \Theta} \sum_{i=1}^n \visible<2->{w_i(\bold{x}) \cdot} \ell(\theta; y_i)
\]

\visible<3->{
\textbf{Weights:}
\begin{eqnarray*}
w^{\text{base}}_i(\bold{x})   & = & 1 \\[0.25cm]
\visible<4->{
w^{\text{tree}}_i(\bold{x})   & = & \sum_{b=1}^B I((\bold{x}_i \in \mathcal{B}_b) \land (\bold{x} \in \mathcal{B}_b)) \\[0.2cm]
\visible<5->{
w^{\text{forest}}_i(\bold{x}) & = & \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t} I((\bold{x}_i \in \mathcal{B}^t_b) \land (\bold{x} \in \mathcal{B}^t_b))
\end{eqnarray*}
}}}
\end{frame}

@@

Schlosser L, Hothorn T, Stauffer R, Zeileis A (2018).
  \dquote{Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.}
  \emph{arXiv 1804.02921}, arXiv.org E-Print Archive.
  \url{http://arxiv.org/abs/1804.02921}
and 
  \emph{Working Papers, Faculty of Economics and Statistics, University of Innsbruck},
  \url{https://EconPapers.repec.org/RePEc:inn:wpaper:2018-08}
